---
title: 'Exercise 2 Data Import & Export - Solutions'
# title: 'Exercise 2 Data Import & Export'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

Now we are in the second exercise session. Non-flat files were promised, so please load the GESIS Panel COVID-19 data.

This first part of the exercises only deals with importing data. Later, in the second exercise, we will turn more to non-flat data files and label them before exporting.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Load the GESIS Panel COVID-19 data.
```

```{block, opts.label = "clues"}
You can either load the SPSS or Stata file for which you need the `haven` package.
```

```{r load_gp, solution = TRUE}
library(haven)

gp_covid <-
  read_spss("../data/ZA5667_v1-1-0.sav")
```

In contrast to the flat files, such as CSV, the variables now have labels. I wonder how the labels are for the first ten variables...

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Print the labels of the first ten variables.
```

```{block, opts.label = "clues"}
You can print labels with the function `sjlabelled::get_label(your_data)`, but you have to make sure only to print the first ten variables.
```

```{r print_labels, solution = TRUE}
library(sjlabelled)

get_label(gp_covid[1:10])
```

Unfortunately, it's all in German. Imagine you are an education researcher, and you are interested in the variable `education_cat`. So you may want to consider translating the variable into English, right?

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Rename the variable label of `age_cat` from "Bildung, kategorisiert" into "Education, categorized".
```

```{block, opts.label = "clues"}
There a two ways: Either using the function `sjlabelled::set_label()` or do it in a pipe with `sjlabelled::var_labels()`.
```

```{r changevarlabel, solution = TRUE}
# either
gp_covid$education_cat <- 
  set_label(
    gp_covid$education_cat, 
    label = "Education, categorized"
  )

# or
library(dplyr)

gp_covid <- 
  gp_covid %>% 
  var_labels(
    education_cat = "Education, categorized"
  )

# proof
get_label(gp_covid$education_cat)
```

Your colleague asks you to provide your new data after changing labels and stuff. Unfortunately, she does not use `R` or `SPSS` and asks you to export your data as a Stata file.

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Export your data as a Stata file.
```

```{block, opts.label = "clues"}
Like in the case of importing, you again need the `haven` package.
```

```{r exportstata, solution = TRUE}
write_stata(gp_covid, "gesis_panel_corona_fancy_panels_final_final.dta")
```