---
title: 'Exercise 2 Data Visualization Part 1 - Solutions'
# title: 'Exercise 2 Data Visualization Part 1'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(gapminder)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

After we have tinkered with the general methods of plotting in R, and before we learn from other plotting methods using `ggplot2`, we now conclude our session with programmaticaly playing with plots. In the previous exercise, we had some repeated tasks to perform. Maybe we can find a more elegant way of plotting multiple plots.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Write a function that automatically creates a table of an input variable and feeds it into a barplot. Moreover, provide an interface to enter the variable name as a character string and the dataset separately by including two commands in the function call.
```

```{block, opts.label = "clues"}
After the function call list two function arguments with `function (variable, data)` and call the variable within the function with `data[[variable]]`.
```

```{r barplotfunction, solution = TRUE}
barplotter <- function (variable, data) {
  barplot(table(data[[variable]]))
}

barplotter("hzcy001a", gp_covid)
```

Pew, that was not about plotting but rather about programming (sorry!). But now you can use this function and include it in loop!

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Take all four variables from the previous exercise session and feed them in a loop which calls the plotting function. You may want again call `par()` before the loop to have all variables nicely plotted in one graph.
```

```{block, opts.label = "clues"}
Define a character vector with the variable names before you build your graph. For the loop, I'd suggest using a for loop which iterates over this character vector.
```

```{r barplotfunctionloop, solution = TRUE}
variable_names <- c("hzcy001a", "hzcy002a", "hzcy003a", "hzcy004a")

par(mfrow = c(2, 2))
for (variable in variable_names) {
  barplotter(variable, gp_covid)
}
```

Before we later start with the other exercises you may want to consider to clean your graphics device with `dev.off()`

```{r cleandevice , message=FALSE}
dev.off()
```