---
title: 'Exercise 1 Confirmatory Analysis - Solutions'
# title: 'Exercise 1 Confirmatory Analysis'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(gapminder)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

We are now entering the phase of doing serious statistics: we run regression models, inspect them, plot and tabulate them.

For this exercise, let's again pretent to be researchers who are interested in the determinants of the subjective likelihood estimation (hoho) of being treated for COVID-19 in a hospital. We suspect that age plays an important role, something we have already assessed by visualizing a similar relationship. So... we need the GESIS Panel COVID-19 data again:

```{r loadgpcovid, message=FALSE, eval=FALSE}
library(dplyr)
library(haven)
library(sjlabelled)

gp_covid <-
  read_sav(
    "../data/gesis_panel_covid19/ZA5667_v1-1-0.sav"
  ) %>% 
  set_na(na = c(-1:-99, 97)) %>% 
  mutate(
    likelihood_hospital = hzcy003a,
    age_cat = as.factor(age_cat)
  ) %>% 
  remove_all_labels()
```

We start our analysis by checking the relationship with an ANOVA. We also include some very, very essential (TM) covariates: gender, political orientation, and marital status. 

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Run an ANOVA on the proposed relationship between subjective likelihood of a treatment in a hospital and age. Also, include all the covariates
```

```{block, opts.label = "clues"}
We had a similar analysis before... you can simply add the control variables with `+`.
```

```{r barplotfunction, solution = TRUE}
serious_anova <- 
  aov(
    likelihood_hospital ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid
  )

summary(serious_anova)
```

