---
title: 'Exercise 2 Confirmatory Analysis - Solutions'
# title: 'Exercise 2 Confirmatory Analysis'
author: 'Johannes Breuer, Stefan Jünger'
date: 'Introduction to R for Data Analysis'
output:
  # unilur::tutorial_html: default
  unilur::tutorial_html_solution: default
---

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
# custom boxes
knitr::opts_template$set(
  clues = list(box.title = "Clues",
               box.body = list(fill = "#fff9dc", colour = "black"),
               box.header = list(fill = "#ffec8b", colour = "black"),
               box.icon = "fa-search",
               box.collapse = TRUE)
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(gapminder)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

In the last exercise, we exploited some regression models for our analysis. That was a good start, but publications need tables of results and also fancy visualizations, e.g., of predictions.

Again, we load our GESIS Panel COVID-19 data and also - just to make sure - re-run the regression models:

```{r loadgpcovid, message=FALSE, eval=FALSE}
library(dplyr)
library(haven)
library(sjlabelled)

gp_covid <-
  read_sav(
    "../data/gesis_panel_covid19/ZA5667_v1-1-0.sav"
  ) %>% 
  set_na(na = c(-1:-99, 97)) %>% 
  mutate(
    likelihood_hospital = hzcy003a,
    age_cat = as.factor(age_cat),
    likelihood_hospital_cut =
      ifelse(
        likelihood_hospital > median(likelihood_hospital, na.rm = TRUE),
        1,
        0
      )
  ) %>% 
  remove_all_labels()

serious_logistic_regression <- 
  glm(
    likelihood_hospital_cut ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid,
    family = binomial(link = "logit")
  )

serious_cauchit_regression <- 
  glm(
    likelihood_hospital_cut ~ age_cat + sex + political_orientation + marstat,
    data = gp_covid,
    family = binomial(link = "cauchit")
  )
```

In the lecture, we started with the visualization of predictions. However, it is good practice to first build a regression table that shows, for example, statistical significance of regression coefficients.

... 

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Build a stargazer regression table of both models to compare them. Use the option `type = "text"` in order to have it printed in the console. Moreover, choose a table style for the journal where you want to publish.
```

```{block, opts.label = "clues"}
You can find an overview of all supported style with `?stargazer::`list of supported styles``.
```

```{r stargazer, solution = TRUE}
library(stargazer)

stargazer(
  serious_logistic_regression,
  serious_cauchit_regression,
  type = "text",
  style = "asr"
)
```

You know what: reviewer 2 ist post-signficant, at least a little. He requested that you should please use confidence intervals at a level of `.93`.

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Rebuild your regression table and add the confidence intervals with the requested level.
```

```{block, opts.label = "clues"}
Have a look at the help page for all possible options `?stargazer?`. Moreover, it may be that CI are still not show, this is due to the custom style.
```

```{r stargazer, solution = TRUE}
library(stargazer)

stargazer(
  serious_logistic_regression,
  serious_cauchit_regression,
  type = "text",
  ci = TRUE,
  ci.level = .93
)
```

The previous exercise might also show that such integrated packages like `stargazer` have limits. Because of these limits, standardized procedures of gathering results like in `broom` are quite attractive as packages like `huxtable` 'understand' this common format.

We'll leave regression tables now and turn to the plotting of the results. Reviewer 2 wants you to that after the third review as you should know...

```{block, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}
Draw a prediction plot of the variable "age_cat" for both models using the `sjPlot` package. Also, try to draw them side by side.
```

```{block, opts.label = "clues"}
You can use the `patchwork` package for creating the joined plot.
```

```{r predictionplot, solution = TRUE}
library(sjPlot)
library(patchwork)

logit_plot <- 
  plot_model(
    serious_logistic_regression,
    type = "pred",
    terms = "age_cat"
    )

cauchit_plot <- 
  plot_model(
    serious_cauchit_regression,
    type = "pred",
    terms = "age_cat"
    )

logit_plot | cauchit_plot
```

Reviewer 2 is happy now. But Reviewer 1, who is more trusted by the editor, thinks estimating predictions based on 'real' average marginal effects (AME) is more cutting-edge. She also thinks that you have proven that choosing a logit or a cauchit link doesn't make a difference. Therefore she'd prefer to see only one plot for the cauchit model as this is not a methodological paper.

```{block, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}
Draw a prediction plot of the variable "age_cat" only for the cauchit model, but this time base on AME.
```

```{block, opts.label = "clues"}
The `margins` package and its `cplot()` function are your friend.
```

```{r predictionplot, solution = TRUE}
library(margins)

cplot(
  serious_cauchit_regression,
  "age_cat",
  what = "prediction"
)
```

Bonus: You may have seen that `cplot()` outputs the actual prediction data automatically. This was annoying in the slides, but you could use this output to draw a plot in `ggplot2`. Moreover, when you toggle the option `draw = FALSE` then only the data are created.

```{block, box.title = "BONUS", box.body = list(fill = "white"), box.icon = "fa-star"}
Draw a prediction plot of the variable "age_cat" only for the cauchit model based on AME in `ggplot2`.
```

```{block, opts.label = "clues"}
1. The column "xvals" is stored as character. You should convert it into numeric.
2. There's a `geom` called `geom_errorbar` which you can use for the confidence intervals
```

```{r predictionplot, solution = TRUE}
library(ggplot2)

prediction_data <-
  cplot(
  serious_cauchit_regression,
  "age_cat",
  what = "prediction",
  draw = FALSE
) %>% 
  dplyr::mutate(
    xvals = as.numeric(xvals)
  )

ggplot(
  prediction_data,
  aes(x = xvals, y = yvals)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = .1) +
  theme_bw()
```